---
title: "Statistical inference project"
author: "Denys Stadnyk"
output:
  pdf_document: default
  html_document: default
---

We are going to analyze ToothGrowth data and perform some Exploratory and Basic Inferential Data Analysis.

ToothGrowth data is a dataset from R datasets package. It gives information about the effect of vitamin C on tooth growth in Guinea pigs.
Basic description of dataset:
The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).

# Exploratory Data Analysis
Load data and get a summary:
```{r}
library(data.table)
ToothGrowth <- data.table(ToothGrowth)
summary(ToothGrowth)
```
Length of odontoblasts varies from 4.2 to 33.9, dose from 0.5 to 2.0. There are only 2 delivery methods (as described above) that are coded with OJ (orange juice) and VC (ascorbic acid), each has 30 observations.

Distribution of tooth length is mostly uniform.
````{r, out.width="300px", fig.align='center'}
hist(ToothGrowth$len, xlab = "Tooth length", main = "Distribution of tooth lengths")
````

What about the relationship of the dose of vitamin C and the length of odontoblasts?
````{r, out.width="350px", fig.align='center'}
boxplot(ToothGrowth$len ~ ToothGrowth$dose, xlab = "Dose", ylab = "Length",
        main = "The relationship between the dose of vitamin and \n odontoblasts length.")
````

From the plot, we clearly see that there is such relationship.
Let's see if we have a similar relationship with both delivery methods.
````{r, out.width="300px", fig.align='center'}
par(mfrow = c(1,2),oma = c(0, 0, 2, 0))
plot(ToothGrowth[supp == "OJ", list(dose, len)], pch = 16, ylim=c(0,36.5),
     main = "Orange juice", ylab = "Length", xlab = "Dose")
abline(lm(ToothGrowth[supp == "OJ", list(dose, len)]$len ~ 
            ToothGrowth[supp == "OJ", list(dose, len)]$dose), lty = 2)

plot(ToothGrowth[supp == "VC", list(dose, len)], pch = 16, ylim=c(0,36.5),
     main = "Ascorbic acid", ylab = "Length", xlab = "Dose")
abline(lm(ToothGrowth[supp == "VC", list(dose, len)]$len ~ 
            ToothGrowth[supp == "VC", list(dose, len)]$dose), lty = 2)
mtext("Odontoblasts length vs vitamine C dose",outer = TRUE, cex = 1.5)
````

# Basic Inferential Data Analysis
From our plots we have seen that probably there is a relationship between the dose and tooth length, also we've seen that efficiency of vitamin C varies in different delivery methods.

## Questions are:
1. Is giving bigger dose makes tooth length bigger?
2. Which delivery method provides a better result?

## Hypothesis 1: Giving bigger dose makes tooth length bigger.
We'll use two sampled t.test.

````{r}
t.test(ToothGrowth[dose == 1]$len, ToothGrowth[dose == 0.5]$len,
       alternative = "greater")$p.value
````
With p-value 6.342e-08 we can say it is a statistically significant result that with dose 1 milligram/day tooth length is more than with 0.5 milligrams/day.
The same test for 2 mg and 1 mg dose:
````{r}
t.test(ToothGrowth[dose == 2]$len, ToothGrowth[dose == 1]$len,
       alternative = "greater")$p.value
````

### Conclusion:
Our hypothesis is true: giving bigger dose makes tooth length bigger.

## Hypothesis 2: Orange juice gives better results than ascorbic acid.
From our previous plots, we can assume that for different doses orange juice and ascorbic acid give different results. So, we test our hypothesis for every dose (0.5, 1, 2).

### Dose 0.5 mg/day
````{r}
t.test(ToothGrowth[dose == 0.5 & supp == "OJ"]$len,
       ToothGrowth[dose == 0.5 & supp == "VC"]$len, alternative = "greater")$p.value
````
With p-value = 0.003179 we can conclude that for 0.5 mg/day dose Orange juice gives better results than ascorbic acid.

### Dose 1 mg/day
````{r}
t.test(ToothGrowth[dose == 1 & supp == "OJ"]$len,
       ToothGrowth[dose == 1 & supp == "VC"]$len, alternative = "greater")$p.value
````
P-value 0.0005192 gives us a statistically significant result of the superiority of orange juice method over ascorbic acid.

### Dose 2 mg/day
Greater hypothesis:
````{r}
t.test(ToothGrowth[dose == 2 & supp == "OJ"]$len,
       ToothGrowth[dose == 2 & supp == "VC"]$len, alternative = "greater")$p.value
````
Not equal hypothesis:
````{r}
t.test(ToothGrowth[dose == 2 & supp == "OJ"]$len,
       ToothGrowth[dose == 2 & supp == "VC"]$len)$p.value
````
As for 2 mg/day dose, we can't say that means are different in the statistically significant way.

### Conclusion:
For 0.5 mg/day and 1 mg/day, our hypothesis is true: orange juice gives better results than ascorbic acid.
For 2 mg/day, there is no statistically significant difference between methods of delivery vitamin C.


# Conclusions and assumptions:
## Assumptions
1. As we do not have a control group, which did not get vitamin C we assume that their tooth length is less than in our data (so there is a practical sense in our analysis).
2. Each population with the same delivery method should follow normal distribution (so we can test it with t.test).

## Conclusions
1. Giving a bigger dose of vitamin C leads to more teeth growth.
2. For the doses of 0.5 and 1 mg/day, it is more efficient to deliver vitamin C with orange juice, for the dose of 2 mg/day there is no statistically significant difference between methods of delivery.